Refinement of Parallel Algorithms down to LLVM

### 2024-03-22 papers/JAR_SI_ITP2022/main.tex
Handled reviewer's comments

\abstract{
  We present a stepwise refinement approach to develop
  verified parallel algorithms, down to efficient LLVM code.
  The resulting algorithms' performance is competitive with their
  counterparts implemented in C/C++.
%
  Our approach is backwards compatible with the Isabelle Refinement Framework,
  such that existing sequential formalizations can easily be adapted
  or re-used.
%
  As case study, we verify a parallel quicksort algorithm that is
  competitive to unverified state-of-the-art algorithms.
}

@@

This paper is an extended version of our ITP 2023 paper~\cite{La22}.
The main new contribution is a verified parallel partitioning algorithm,
which significantly improves the efficiency and scalability of our sorting algorithm from~\cite{La22}.
To this end, we added a description of the interval list data structure (Section~\ref{sec:mod_ds})
and the \is{with_idxs} combinator (Section~\ref{sec:array_split}), which are required by the parallel partitioner.
The parallel partitioning algorithm itself is described in Section~\ref{sec:ppar}, and Section~\ref{sec:benchmark}
contains updated and more extensive benchmarks.

@@

\noindent The formalization is available at \href{https://www21.in.tum.de/~lammich/isabelle_llvm_par/}{www21.in.tum.de/$\sim$lammich/isabelle\_llvm\_par}.

@@

On top of the back-end, we formalize a concurrent separation logic~\cite{OH04} and implement a verification condition generator (VCG). (Section~\ref{sec:seplogic}).

@@

We mainly use Isabelle/HOL notation, with some shortcuts and adaptations for presentation in a paper.
In this section, we give examples of the more unusual notations: for implication, we always write \is{==>},
e.g., \is{\<forall>x. x<4 ==> x<=4} (in Isabelle, one must use \is{-->} here). Free variables are
universally quantified at the top-level.

Type variables are \is{'a, 'b, ...} and function types are written curried as \is{'a => 'b => 'c}.
Types can be annotated to any term, e.g., \is{(x::nat) < 4 ==> x<=3}.
Function application is written as \is{f x y}, and function update is \is{f(x:=y)}.
For definitions, we use \is{==}, e.g., \is{f x == x+1}.

Algebraic datatypes are written as \is{'a option == Some (the: 'a) | None}. This also defines the selector function \is{the :: 'a option => 'a}.
Values of the tuple type \is{'a_1 \x ... \x 'a$_n$} are written as \is{(a_1,...,a$_n$)}. Lists have type \is{'a list},
and we write \is{[]} for the empty list, \is{x#xs} for the list with head \is{x} and tail \is{xs}, and \is{xs_1@xs_2} for list concatenation.
We also use the list notation \is{[1,2,3]}. The length of list \is{xs} is \is{|xs|}.

The empty set is \is${} :: 'a set$. The set \is${l..<h}$ contains all elements between $l$ inclusive and $h$ exclusive, and the set \is${l..}$ contains all elements greater than or equal to $l$.
Disjoint union is written as \is{$\dot\cup$}. We only write \is{$\dot\cup$} if it is clear from the context where the disjointness constraint belongs to.
Otherwise, we will explicitly write, e.g., \is$ s = s_1 \union s_2  &  s_1 \inter s_2 = {}$. The cardinality of the finite set \is{s} is \is{|s|}.

Lambda abstraction is written as \is{\<lambda>x. x+1}. When clear from the context, we omit the \is{\<lambda>}, e.g. \is{spec x. x>42}.
We also use Haskell-like sections for infix operators, e.g. \is{(+1)} for \is{\<lambda>x. x+1}.
We also use underscores to indicate the parameter positions, e.g., \is${_..<_}$ for \is$\<lambda>l h. {l..<h}$.

@@

  Now assume that we have a state (memory) type \is{\mu}, and an access report type \is{\rho}, which forms a monoid (\is{0,+}).
  With this, we define our state-nondeterminism-error monad with access reports, just called \is{M} for brevity:
  \begin{lstlisting}
    'x M == \mu => ('x \x \rho \x \mu) neM
    return$_M$ x \mu == return$_{ne}$ (x,0,\mu)
    bind$_M$ m f \mu == (x_1,r_1,\mu) <- m \mu; (x_2,r_2,\mu) <- f x_1 \mu; return$_{ne}$ (x_2,r_1+r_2,\mu)
  \end{lstlisting}
  Here, \is{return} does not change the state, and reports no accesses (\is$0$),
  and \is{bind} sequentially composes the executions, threading through the state $\mu$ and adding up the access reports \is$r_1$ and \is$r_2$.

@@

  \subsection{Memory Model}
  Our memory model supports blocks of values, where values can be integers, structures, or pointers into a block:
  \begin{lstlisting}
    datatype addr == ADDR (bidx: nat) (idx: nat)
    datatype ptr == PTR_NULL   |   PTR_ADDR (the_addr: addr)
    datatype val == LL_INT lint   |   LL_STRUCT val list   |   LL_PTR ptr

    datatype block == FRESH   |   FREED   |   is_alloc: ALLOC (vals: "val list")
    typedef memory == { \mu :: nat => block. finite {b. \mu b ~= FRESH} }
  \end{lstlisting}
  A block is either fresh, freed, or allocated, and a memory is a mapping from block indexes to blocks,
  such that only finitely many blocks are not fresh. Every block's state transitions from fresh to
  allocated to freed. This avoids ever reusing the same block, and thus allows us to semantically detect use after free errors.
  Every program execution can only allocate finitely many blocks,
  such that we will never run out of fresh blocks\footnote{If the actual system does run out of memory, we will terminate the program in a defined way.}.
  An allocated block contains an array of values, modelled as a list. Thus, an address consists of a block number, and an index into the array.

  To access and modify memory, we define the functions \is{valid}, \is{get}, and \is{put}:
  \begin{lstlisting}
    valid \mu (ADDR b i) == is_alloc (\mu b) & i<|vals (\mu b)|
    get \mu (ADDR b i) == vals (\mu b) ! i
    put \mu (ADDR b i) x == \mu(b := ALLOC ((vals (\mu b))[i:=x]))
  \end{lstlisting}
  where \is{|xs|} is the length of list \is{xs}, \is{xs!i} returns the \is$i$th element of list \is$xs$,
  and \is{xs[i:=x]} replaces the \is$i$th element of \is{xs} by \is{x}.

@@

  \subsection{Access Reports}
  We now fix the state of the M-monad to be memory, and the access reports to be tuples \is{(r,w,a,f)} of
  read and written addresses, as well as sets of allocated and freed blocks:
  \begin{lstlisting}
    acc == addr set \x addr set \x nat set \x nat set
    0 == ( {},{},{},{} )
    (r_1,w_1,a_1,f_1) + (r_2,w_2,a_2,f_2) == ( r_1\union r_2, w_1\union w_2, a_1\union a_2, f_1\union f_2 )
  \end{lstlisting}

@@

    norace (r_1,w_1,a_1,f_1) (r_2,w_2,a_2,f_2) ==
      let m_1 = w_1 \<union> { ADDR b i. b \<in> a_1 \<union> f_1 } in
      let m_2 = w_2 \<union> { ADDR b i. b \<in> a_2 \<union> f_2 } in
        (r_1 \<union> m_1) \<inter> m_2 = {}   &   m_1 \<inter> (r_2 \<union> m_2) = {}
  \end{lstlisting}
  The combine function joins the access reports and memories,
  preferring allocated over fresh, and freed over allocated memory. When joining two allocated blocks, the written addresses
  from the access report are used to join the blocks. We skip the rather technical definition of combine, and just state the
  relevant properties: Let \is{\rho_1=(r_1,w_1,a_1,f_1)} and \is{\rho_2=(r_2,w_2,a_2,f_2)} be feasible and race free access reports,
  and \is{\mu_1, \mu_2} be memories that have evolved from a common memory \is{\mu}, consistently with
  the access reports \is{\rho_1, \rho_2}. Let \is{(\rho',\mu') = combine (\rho_1,\mu_1) (\rho_2,\mu_2)}. Then
  \begin{lstlisting}
    (1)  \mu' b = FRESH <--> \mu b = FRESH \and b ~: a_1 \union a_2
    (2)  is_alloc (\mu' b) <--> (is_alloc (\mu b) \or b \<in> a_1 \union a_2) \and b ~: f_1 \union f_2
    (3)  \mu' b = FREED <--> \mu b = FREED \or b \in f_1 \union f_2
  \end{lstlisting}
  Moreover, for all addresses \is{addr = ADDR b i} with \is{valid \mu' addr}:
  \begin{lstlisting}
    (4)  addr \in w_1 \or b \in a_1 ==> get \mu' addr = get \mu_1 addr
    (5)  addr \in w_2 \or b \in a_2 ==> get \mu' addr = get \mu_2 addr
    (6)  addr ~: w_1\<union>w_2 & b ~: a_1\<union>a_2 ==> get \mu' addr = get \mu addr
  \end{lstlisting}
  The properties (1)--(3) define the state of blocks in the combined memory:
  a fresh block in \is{\mu'} was fresh already in \is{\mu}, and has not been allocated (1);
  an allocated block was already allocated or has been allocated, but has not been freed (2); and a freed block was already freed, or has been freed (3).
  The properties (4)--(6) define the content: addresses written or allocated in the first or second execution get their content from \is{\mu_1} (4) or \is{\mu_2} (5) respectively.
  Addresses not written nor allocated at all keep their original content (6).

@@

  \subsection{LLVM Instructions}
  Based on the M-monad, we define shallowly embedded LLVM instructions. For most instructions,
  this is analogous to the sequential case~\cite{La19-llvm}.
  Additionally, we define an instruction for parallel function call:
  \begin{lstlisting}
    llc_par f g a b == f a || g b
  \end{lstlisting}
  The code generator only accepts this, if \is{f} and \is{g} are constants (i.e., function names).
  It then generates some type-casting boilerplate, and a call to an external
  \is{parallel} function, which we implement using the Threading Building Blocks~\cite{TBB} library:
  \begin{lstlisting}[language=C++]
  void parallel(void (*f1)(void*), void (*f2)(void*), void *x1, void *x2) {
    tbb::parallel_invoke([=]{f1(x1);}, [=]{f2(x2);}); }
  \end{lstlisting}
  I.e., the two functions \is{f1(x1)} and \is{f2(x2)} are called in parallel. The generated boilerplate code sets up \is{x1} and \is{x2} to point to both, the actual arguments and space for the results.

@@

  For our memory model, we define the following abstraction function:
  \begin{lstlisting}
    \alpha :: memory => (addr => val option) \x (nat => nat option)
    \alpha \mu == (\alpha$_m$ \mu, \alpha$_b$ \mu)

@@

    disjoint (r,w,a,f) (m,b) == (\<forall>addr. m addr ~= 0 ==> addr ~: r \union w & addr.bidx ~: f)
                                 & (\<forall>i. b i ~= 0 ==> i ~: f)
  \end{lstlisting}
  that is, when executed on memory \is{\mu}, the program \is{c} does not fail, every return value \is{x} and new memory \is{\mu'} satisfies \is{Q},
  and no memory described by the frame \is{amf} is accessed.

@@

  \begin{example}[Swapping multiple elements]\label{ex:swap_spec}
    We specify an operation to perform multiple swaps. It takes two disjoint sets of indices \is{s_1} and \is{s_2}, and a list \is{xs}. It then swaps each index in \is{s_1} with some index in \is{s_2}.
    The precondition of this operation assumes that the index sets are in range, disjoint, and have the same cardinality:
    \begin{lstlisting}
      swap_spec_pre s_1 s_2 xs == s_1 \inter s_2 = {} & s_1 \union s_2 \<subseteq> {0..<|xs|} & |s_1| = |s_2|
    \end{lstlisting}
    The postcondition ensures that the resulting list is a permutation of the original list,
    the elements at indexes outside \is{s_1 \union s_2} are unchanged, and that each element in \is{s_1} is swapped with one in \is{s_2}:
    \begin{lstlisting}
      swap_spec_post s_1 s_2 xs xs' ==
        mset xs' = mset xs & (\<forall>i\<notin>s_1 \union s_2. i < |xs| ==> xs'!i=xs!i)
      & (\<forall>i\<in>s_1. \<exists>j\<in>s_2. xs'!i=xs!j) & (\<forall>j\<in>s_2. \<exists>i\<in>s_1. xs'!j=xs!i)
    \end{lstlisting}

@@

  \begin{example}(Refinement of lists to arrays)\label{ex:ref_array}
    We define abstract programs for indexing and updating a list:
    \begin{lstlisting}
      lget xs i == assert (i<|xs|); return xs!i
      lset xs i x == assert (i<|xs|); return xs[i:=x]
    \end{lstlisting}
    These programs assert that the index is in bounds, and then return the accessed element (\is{xs!i})
    or the updated list (\is{xs[i:=x]}) respectively. The following assertion links a pointer to a list of
    elements stored at the pointed-to location:
    \begin{lstlisting}
      arr$_A$ p xs = range {0..<|xs|} p (%i. xs!i)
    \end{lstlisting}
    That is, for every $i<|xs|$, $p+i$ points to the $i$th element of $xs$.
    Assertions like \is{arr$_A$}, that relate concrete to abstract values, are called \emph{refinement relations}.
    If we want to emphasize that they depend on the heap, we also call them refinement \emph{assertions}.

    Indexing and updating of arrays is implemented by:
    \begin{lstlisting}
      aget p i == ll_ofs_ptr p i; ll_load p
      aset p i x == ll_ofs_ptr p i; ll_store x p; return p
    \end{lstlisting}
    The abstract and concrete programs are linked by the following refinement theorems:
    \begin{lstlisting}
      hnr (arr$_A$ xs$_\dagger$ xs * idx$_A$ i$_\dagger$ i)$\hspace*{-1pt}$ (aget xs$_\dagger$ i$_\dagger$)$\hspace*{-1pt}$ (arr$_A$ xs$_\dagger$ xs$\hspace*{1pt}$*$\hspace*{1pt}$idx$_A$ i$_\dagger$ i)$\hspace*{-1pt}$ id$_A$$\hspace*{-1pt}$ $\top$$\hspace*{-1pt}$ (lget xs i)
      hnr (arr$_A$ xs$_\dagger$ xs * idx$_A$ i$_\dagger$ i) (aset xs$_\dagger$ i$_\dagger$ x) (idx$_A$ i$_\dagger$ i) arr$_A$ (%r. r=xs$_\dagger$) (lset xs i x)
    \end{lstlisting}
    That is, if the list \is$xs$ is refined by array \is{xs$_\dagger$}, and the natural number \is$i$ is refined
    by the fixed-width\footnote{We use Isabelle's word library here, which encodes the actual width as a type variable,
    such that our functions work with any bit width. For code generation, we will fix the width to 64 bit.} word \is{i$_\dagger$} (\is{idx$_A$ i$_\dagger$ i}),
    the \is{aget} operation will return the same result as the \is{lget} operation (\is{id$_A$}).
    The resulting memory will still contain the original array. Note that there is no explicit precondition that the array access
    is in bounds, as this follows already from the assertion in the abstract \is{lget} operation.
    The \is{aset} operation will return a pointer to an array that refines the updated list returned by \is{lset}.
    As the array is updated in place, the original refinement of the array is no longer valid.
    Moreover, the returned pointer \is{r} will be the same as the argument pointer \is{xs$_\dagger$}.
    This information is important for refining to parallel programs on disjoint parts of an
    array (cf.\ Section~\ref{sec:array_split}).

    To increase readability, we introduce an (almost) point-free notation for refinement theorems.
    The theorems for the array operations above can also be written as:
    \begin{lstlisting}
      aget, lget : arr$_A$ -> idx$_A$ -> id$_A$
      aset, lset : arr$_A$^d:xs$_\dagger$ -> idx$_A$ -> id$_A$ -> arr$_A$:r [r=xs$_\dagger$]
    \end{lstlisting}
    The first theorem simply states that the first argument is refined by \is{arr$_A$},
    the second argument by \is{idx$_A$}, and the result by \is{id$_A$}.
    The second theorem adds the annotation $\cdot^d$ to the refinement for the array argument,
    indicating that this argument will be destroyed, i.e., the refinement is no longer valid when the function returns.
    Moreover, it binds the array argument to the name \is{xs$_\dagger$} and the result to $r$. These names are used in the pointer equality predicate \is{[r=xs$_\dagger$]}
    at the end, indicating that the result will be the same pointer as the array argument.

@@

    iv l h == (l,h)                                                 iv, {_..<_} : id$_R$ -> id$_R$ -> iv$_R$
    iv_is_empty (l,h) == h<=l                                    iv_is_empty, (={}) : iv$_R$ -> id$_R$
    iv_inter (l_1,h_1) (l_2,h_2) == ( max l_1 l_2, min h_1 h_2 )     iv_inter, (\inter) : iv$_R$ -> iv$_R$ -> iv$_R$
    iv_card (l,h) == if h<l then 0 else h-l                iv_card, |_| : iv$_R$ -> id$_R$
  \end{lstlisting}
  Note that \is$h<l ==> {l..<h} = {}$, and we do not enforce $l\le h$ for our representation. Thus, no checks are needed for construction and intersection.
  However, we use a check to avoid underflow when computing the cardinality.

@@

    ivl_empty == []                                 ivl_empty, {} : ivl$_R$
    ivl_is_empty ls == ls=[]                       ivl_is_empty, (={}) : ivl$_R$ -> id$_R$
    ivl_dj_un s ls == s#ls                         ivl_dj_un, ($\dot\cup$) : id$_R$ -> ivl$_R$ -> ivl$_R$
    ivl_card ls == fold (%s c. c=c+|s|) ls 0     ivl_card, card : ivl$_R$ -> id$_R$
    ivl_split (s#ls) == (s,ls)                      ivl_split, split : ivl$_R$ -> id$_R$ \x ivl$_R$
      split s == assert (s~={}); spec (s_1,s_2). s = s_1 $\dot\cup$ s_2 & s_1~={}
  \end{lstlisting}

>>>>

\abstract{
  We present a stepwise refinement approach to develop
  verified parallel algorithms, down to efficient LLVM code.
  The resulting algorithms' performance is competitive with their
  counterparts implemented in C++.
%
  Our approach is backwards compatible with the Isabelle Refinement Framework,
  such that existing sequential formalizations can easily be adapted
  or re-used.
%
  As case study, we verify a parallel quicksort algorithm that is
  competitive to unverified state-of-the-art algorithms.
}

@@

This paper is an extended version of our ITP 2022 paper~\cite{La22}.
The main new contribution is a verified parallel partitioning algorithm,
which significantly improves the efficiency and scalability of our sorting algorithm from~\cite{La22}.
To this end, we added a description of the interval list data structure (Section~\ref{sec:mod_ds})
and the \is{with_idxs} combinator (Section~\ref{sec:array_split}), which are required by the parallel partitioner.
The parallel partitioning algorithm itself is described in Section~\ref{sec:ppar}, and Section~\ref{sec:benchmark}
contains updated and more extensive benchmarks.

@@

The formalization is available at \href{https://www21.in.tum.de/~lammich/isabelle_llvm_par/}{www21.in.tum.de/$\sim$lammich/isabelle\_llvm\_par}.

@@

On top of the back-end, we formalize a concurrent separation logic~\cite{OH04} and implement a verification condition generator (VCG), cf.\ Section~\ref{sec:seplogic}.

@@

We mainly use Isabelle/HOL notation, with some shortcuts and adaptations for presentation in a paper.
In this section, we give examples of the more unusual notations: for implication, we always write \is{==>},
e.g., \is{\<forall>x. x\:<\:4 ==> x\:<=\:4} (in Isabelle, one must use \is{-->} here). Free variables are
universally quantified at the top-level.

Type variables are \is{'a, 'b, ...} and function types are written curried as \is{'a => 'b => 'c}.
Types can be annotated to any term, e.g., \is{(x::nat)\:<\:4 ==> x\:<=\:3}.
Function application is written as \is{f x y}, and function update is \is{f(x:=y)}.
For definitions, we use \is{==}, e.g., \is{f x == x+1}.

Algebraic datatypes are written as \is{'a option == Some (the: 'a) | None}. This also defines the selector function \is{the :: 'a option => 'a}.
Values of the tuple type \is{'a_1 \x ... \x 'a$_n$} are written as \is{(a_1,...,a$_n$)}. Lists have type \is{'a list},
and we write \is{[]} for the empty list, \is{x#xs} for the list with head \is{x} and tail \is{xs}, and \is{xs_1@\:xs_2} for list concatenation.
We also use the list notation \is{[1,2,3]}. The length of list \is{xs} is \is{|xs|}.

The empty set is \is${} :: 'a set$. The set \is${l..<h}$ contains all elements between $l$ inclusive and $h$ exclusive, and the set \is${l..}$ contains all elements greater than or equal to $l$.
Disjoint union is written as \is{$\dot\cup$}. We only write \is{$\dot\cup$} if it is clear from the context where the disjointness constraint belongs to.
Otherwise, we will explicitly write, e.g., \is$ s = s_1 \union s_2  &  s_1 \inter s_2 = {}$. The cardinality of a finite set \is{s} is \is{|s|}.

Lambda abstraction is written as \is{\<lambda>x. x+1}. When clear from the context, we omit the \is{\<lambda>}, e.g. \is{spec x. x\:>\:42}.
We also use Haskell-like sections for infix operators, e.g. \is{(+1)} for \is{\<lambda>x. x+1}.
We also use underscores to indicate the parameter positions, e.g., \is${_..<_}$ for \is$\<lambda>l h. {l..<h}$.

@@

  Now assume that we have a state (memory) type \is{\mu}, and an access report type \is{\rho}, which forms a monoid (\is{0,+}).
  With this, we define our state-nondeterminism-error monad with access reports, just called \is{M} for brevity:
  \begin{lstlisting}
    'x M == \mu => ('x \x \rho \x \mu) neM
    return$_M$ x \mu == return$_{ne}$ (x,0,\mu)
    bind$_M$ m f \mu == (x_1,r_1,\mu) <- m \mu; (x_2,r_2,\mu) <- f x_1 \mu; return$_{ne}$ (x_2,r_1+r_2,\mu)
  \end{lstlisting}
  Here, \is{return} does not change the state, and reports no accesses (\is$0$),
  and \is{bind} sequentially composes the executions, threading through the state $\mu$ and adding up the access reports \is$r_1$ and \is$r_2$. Note that we use the names \is$r :: \rho$ for reports, \is$\mu :: \mu$ for memory, and \is$m :: _ M$ for monads.

@@

  \subsection{Memory Model}
  Our memory model supports blocks of values, where values can be integers, structures, or pointers into a block:
  \begin{lstlisting}
    bidx == nat
    addr == ADDR bidx nat
    ptr == PTR_NULL   |   PTR_ADDR (the_addr: addr)
    val == LL_INT lint   |   LL_STRUCT val list   |   LL_PTR ptr

    block == FRESH   |   FREED   |   is_alloc: ALLOC (vals: "val list")
    typedef memory == { \mu :: bidx => block. finite {b. \mu b ~= FRESH} }
  \end{lstlisting}
  A block is either fresh, freed, or allocated, and a memory is a mapping from block indexes (\is{bidx}) to blocks,
  such that only finitely many blocks are not fresh. Every block's state transitions from fresh to
  allocated to freed. This avoids ever reusing the same block, and thus allows us to semantically detect use after free errors.
  Every program execution can only allocate finitely many blocks,
  such that we will never run out of fresh blocks\footnote{If the actual system does run out of memory, we will terminate the program in a defined way.}.
  An allocated block contains an array of values, modelled as a list. Thus, an address \is$ADDR bi i$ consists of a block index \is$bi$, and an index \is{i} into the array.

  To access and modify memory, we define the functions \is{valid}, \is{get}, and \is{put}:
  \begin{lstlisting}
    valid \mu (ADDR b i) == is_alloc (\mu b) & i < |vals (\mu b)|
    get \mu (ADDR b i) == vals (\mu b) ! i
    put \mu (ADDR b i) x == \mu(b := ALLOC ((vals (\mu b))[i:=x]))
  \end{lstlisting}
  where \is{|xs|} is the length of list \is{xs}, \is{xs!i} returns the \is$i$th element of list \is$xs$,
  and \is{xs[i:=x]} replaces the \is$i$th element of \is{xs} by \is{x}.

@@

  \subsection{Access Reports}
  We now fix the state of the M-monad to be memory, and the access reports to be tuples \is{(r,w,a,f)} of
  read and written addresses, as well as sets of allocated and freed blocks:
  \begin{lstlisting}
    acc == addr set \x addr set \x bidx set \x bidx set
    0 == ( {},{},{},{} )
    (r_1,w_1,a_1,f_1) + (r_2,w_2,a_2,f_2) == ( r_1 \union r_2, w_1 \union w_2, a_1 \union a_2, f_1 \union f_2 )
  \end{lstlisting}

@@

    norace (r_1,w_1,a_1,f_1) (r_2,w_2,a_2,f_2) ==
      let m_1 = w_1 \<union> { ADDR b i. b \<in> a_1 \<union> f_1 } in
      let m_2 = w_2 \<union> { ADDR b i. b \<in> a_2 \<union> f_2 } in
        (r_1 \<union> m_1) \<inter> m_2 = {}   &   m_1 \<inter> (r_2 \<union> m_2) = {}
  \end{lstlisting}
  The combine function joins the access reports and memories,
  preferring allocated over fresh, and freed over allocated memory. When joining two allocated blocks, the written addresses
  from the access report are used to join the blocks. We skip the rather technical definition of combine, and just state the
  relevant properties: Let \is{\rho_1=(r_1,w_1,a_1,f_1)} and \is{\rho_2=(r_2,w_2,a_2,f_2)} be feasible and race free access reports,
  and \is{\mu_1, \mu_2} be memories that have evolved from a common memory \is{\mu}, consistently with
  the access reports \is{\rho_1, \rho_2}. Let \is{(\rho',\mu') = combine (\rho_1,\mu_1) (\rho_2,\mu_2)}. Then
  \begin{lstlisting}
    (1)  \mu' b = FRESH <--> \mu b = FRESH \and b ~: a_1 \union a_2
    (2)  is_alloc (\mu' b) <--> (is_alloc (\mu b) \or b \<in> a_1 \union a_2) \and b ~: f_1 \union f_2
    (3)  \mu' b = FREED <--> \mu b = FREED \or b \in f_1 \union f_2
  \end{lstlisting}
  Moreover, for all addresses \is{addr = ADDR b i} with \is{valid \mu' addr}:
  \begin{lstlisting}
    (4)  addr \in w_1 \or b \in a_1 ==> get \mu' addr = get \mu_1 addr
    (5)  addr \in w_2 \or b \in a_2 ==> get \mu' addr = get \mu_2 addr
    (6)  addr ~: w_1\:\<union>\:w_2 & b ~: a_1\:\<union>\:a_2 ==> get \mu' addr = get \mu addr
  \end{lstlisting}
  The properties (1)--(3) define the state of blocks in the combined memory:
  a fresh block in \is{\mu'} was fresh already in \is{\mu}, and has not been allocated (1);
  an allocated block was already allocated or has been allocated, but has not been freed (2); and a freed block was already freed, or has been freed (3).
  The properties (4)--(6) define the content: addresses written or allocated in the first or second execution get their content from \is{\mu_1} (4) or \is{\mu_2} (5) respectively.
  Addresses not written nor allocated at all keep their original content (6).

@@

  \subsection{LLVM Instructions}
  Based on the M-monad, we define shallowly embedded LLVM instructions. For most instructions,
  this is analogous to the sequential case~\cite{La19-llvm}.
  Additionally, we define an instruction for a parallel function call:
  \begin{lstlisting}
    llc_par f g a b == f a || g b
  \end{lstlisting}
  The code generator only accepts this, if \is{f} and \is{g} are constants (i.e., function names).
  It then generates some type-casting boilerplate, and a call to an external
  \is{parallel} function, which we implement using the Threading Building Blocks~\cite{TBB} library:
  \begin{lstlisting}[language=C++, basicstyle=\ttfamily]
  void parallel(void (*f1)(void*), void (*f2)(void*), void *x1, void *x2) {
    tbb::parallel_invoke([=]{f1(x1);}, [=]{f2(x2);}); }
  \end{lstlisting}
  I.e., the two functions \is{f1(x1)} and \is{f2(x2)} are called in parallel. The generated boilerplate code sets up \is{x1} and \is{x2} to point to both, the actual arguments and space for the results.

@@

  For our memory model, we define the following abstraction function:
  \begin{lstlisting}
    \alpha :: memory => (addr => val option) \x (bidx => nat option)
    \alpha \mu == (\alpha$_m$ \mu, \alpha$_b$ \mu)

@@

    disjoint (r,w,a,f) (m,b) == (\<forall>addr. m addr ~= 0 ==> addr ~: r \union w & addr.bidx ~: f)
                                 & (\<forall>i. b i ~= 0 ==> i ~: f)
  \end{lstlisting}
  that is, when executed on memory \is{\mu}, the program \is{c} does not fail, every return value \is{x} and new memory \is{\mu'} satisfies \is{Q},
  and no memory described by the frame \is{amf}~~is accessed.

@@

  \begin{example}[Swapping multiple elements]\label{ex:swap_spec}
    We specify an operation to perform multiple swaps. It takes two disjoint sets of indexes \is{s_1} and \is{s_2}, and a list \is{xs}. It then swaps each index in \is{s_1} with some index in \is{s_2}.
    The precondition of this operation assumes that the index sets are in range, disjoint, and have the same cardinality:
    \begin{lstlisting}
      swap_spec_pre s_1 s_2 xs == s_1 \inter s_2 = {} & s_1 \union s_2 \<subseteq> {0..<|xs|} & |s_1| = |s_2|
    \end{lstlisting}
    The postcondition ensures that the resulting list is a permutation of the original list,
    the elements at indexes outside \is{s_1 \union s_2} are unchanged, and that each element in \is{s_1} is swapped with one in \is{s_2}:
    \begin{lstlisting}
      swap_spec_post s_1 s_2 xs xs' ==
        mset xs' = mset xs & (\<forall>i\<notin>s_1 \union s_2. i < |xs| ==> xs'!i = xs!i)
      & (\<forall>i\<in>s_1. \<exists>j\<in>s_2. xs'!i = xs!j) & (\<forall>j\<in>s_2. \<exists>i\<in>s_1. xs'!j = xs!i)
    \end{lstlisting}
    Here, \is{mset xs} is the multiset of elements of the list \is{xs}, and \is{mset xs = mset xs'} is the standard way to express permutation in Isabelle.

@@

  \begin{example}(Refinement of lists to arrays)\label{ex:ref_array}
    We define abstract programs for indexing and updating a list:
    \begin{lstlisting}
      lget xs i == assert (i < |xs|); return xs!i
      lset xs i x == assert (i < |xs|); return xs[i:=x]
    \end{lstlisting}
    These programs assert that the index is in bounds, and then return the accessed element (\is{xs!i})
    or the updated list (\is{xs[i:=x]}) respectively. The following assertion links a pointer to a list of
    elements stored at the pointed-to location:
    \begin{lstlisting}
      arr$_A$ p xs = range {0..<|xs|} p (%i. xs!i)
    \end{lstlisting}
    That is, for every $i<|xs|$, $p+i$ points to the $i$th element of $xs$.
    Assertions like \is{arr$_A$}, that relate concrete to abstract values, are called \emph{refinement relations}.
    If we want to emphasize that they depend on the heap, we also call them refinement \emph{assertions}.

    Indexing and updating of arrays is implemented by:
    \begin{lstlisting}
      aget p i == ll_ofs_ptr p i; ll_load p
      aset p i x == ll_ofs_ptr p i; ll_store x p; return p
    \end{lstlisting}
    where \is{ll_ofs_ptr} is the Isabelle-LLVM instruction for offsetting a pointer by an index.
    The abstract and concrete programs are linked by the following refinement theorems:
    \begin{lstlisting}
      hnr (arr$_A$ xs$_\dagger$ xs * idx$_A$ i$_\dagger$ i)$\hspace*{-1pt}$ (aget xs$_\dagger$ i$_\dagger$)$\hspace*{-1pt}$ (arr$_A$ xs$_\dagger$ xs$\hspace*{1pt}$*$\hspace*{1pt}$idx$_A$ i$_\dagger$ i)$\hspace*{-1pt}$ id$_A$$\hspace*{-1pt}$ $\top$$\hspace*{-1pt}$ (lget xs i)
      hnr (arr$_A$ xs$_\dagger$ xs * idx$_A$ i$_\dagger$ i) (aset xs$_\dagger$ i$_\dagger$ x) (idx$_A$ i$_\dagger$ i) arr$_A$ (%r. r\:=\:xs$_\dagger$) (lset xs i x)
    \end{lstlisting}
    That is, if the list \is$xs$ is refined by array \is{xs$_\dagger$}, and the natural number \is$i$ is refined
    by the fixed-width\footnote{We use Isabelle's word library here, which encodes the actual width as a type variable,
    such that our functions work with any bit width. For code generation, we will fix the width to 64 bit.} word \is{i$_\dagger$} (\is{idx$_A$ i$_\dagger$ i}),
    the \is{aget} operation will return the same result as the \is{lget} operation (\is{id$_A$}).
    The resulting memory will still contain the original array. Note that there is no explicit precondition that the array access
    is in bounds, as this follows already from the assertion in the abstract \is{lget} operation.
    The \is{aset} operation will return a pointer to an array that refines the updated list returned by \is{lset}.
    As the array is updated in place, the original refinement of the array is no longer valid.
    Moreover, the returned pointer \is{r} will be the same as the argument pointer \is{xs$_\dagger$}.
    This information is important for refining to parallel programs on disjoint parts of an
    array (cf.\ Section~\ref{sec:array_split}).

    To increase readability, we introduce an (almost) point-free notation for refinement theorems.
    The theorems for the array operations above can also be written as:
    \begin{lstlisting}
      aget, lget : arr$_A$ -> idx$_A$ -> id$_A$
      aset, lset : arr$_A$^d:xs$_\dagger$ -> idx$_A$ -> id$_A$ -> arr$_A$:r [r\:=\:xs$_\dagger$]
    \end{lstlisting}
    The first theorem simply states that the first argument is refined by \is{arr$_A$},
    the second argument by \is{idx$_A$}, and the result by \is{id$_A$}.
    The second theorem adds the annotation $\cdot^d$ to the refinement for the array argument,
    indicating that this argument will be destroyed, i.e., the refinement is no longer valid when the function returns.
    Moreover, it binds the array argument to the name \is{xs$_\dagger$} and the result to $r$. These names are used in the pointer equality predicate \is{[r\:=\:xs$_\dagger$]}
    at the end, indicating that the result will be the same pointer as the array argument.

@@

    iv l h == (l,h)                                                 iv, {_..<_} : id$_R$ -> id$_R$ -> iv$_R$
    iv_is_empty (l,h) == h\:<=\:l                                   iv_is_empty, (={}) : iv$_R$ -> id$_R$
    iv_inter (l_1,h_1) (l_2,h_2) == ( max l_1 l_2, min h_1 h_2 )     iv_inter, (\inter) : iv$_R$ -> iv$_R$ -> iv$_R$
    iv_card (l,h) == if h\:<\:l then 0 else h-l               iv_card, |_| : iv$_R$ -> id$_R$
  \end{lstlisting}
  Note that \is$h<l ==> {l..<h} = {}$, and we do not enforce $l\le h$ for our representation. Thus, no checks are needed for construction and intersection.
  However, we use a check to avoid underflow when computing the cardinality.

@@

    ivl_empty == []                                 ivl_empty, {} : ivl$_R$
    ivl_is_empty ls == ls = []                     ivl_is_empty, (={}) : ivl$_R$ -> id$_R$
    ivl_dj_un s ls == s#ls                         ivl_dj_un, ($\dot\cup$) : id$_R$ -> ivl$_R$ -> ivl$_R$
    ivl_card ls == fold (%s c. c=c+|s|) ls 0     ivl_card, card : ivl$_R$ -> id$_R$
    ivl_split (s#ls) == (s,ls)                      ivl_split, split : ivl$_R$ -> id$_R$ \x ivl$_R$
      split s == assert (s ~= {}); spec (s_1,s_2). s = s_1 $\dot\cup$ s_2 & s_1 ~= {}
  \end{lstlisting}
